---
title: "FYP_pilot_goalRating"
author: "Bernice Cheung"
date: "1/24/2020"
output: html_document
---
# Set up the environment
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, results = 'hide', warning=FALSE}
library(tidyverse)
library(psych)
library(ggplot2)
library(knitr)
library(corrplot)
library(lme4)
library(nFactors)
library(GPArotation)
```

Load cleaned dataset, which contains 48 subjects and all variables related to goal ratings. The long format is gathered by dimension and has renamed dimensions by their names and numbered the goal within each participant.
```{r}
# load cleaned dataset for goal ratings
goalDf <- read.csv("./input/goalRating.csv")
goalDf_long <- read.csv("./input/goalRating_long_R.csv")
indivDiffDf <- read.csv("./input/indivDiffDf.csv")
```

# The number of listed goals
participants on average listed 3.75 goals. By comparing the number of goals they listed to the number of goals they claimed to have: 12% participants have more than 5 goals; 4% participants have 1 goal; 10% participants listed one more goal than they originally estimated; 10 participants (not including those who have more than 5 goals) listed fewer goals than they originally estimated.   

The goals participants listed might be biased by the task, but it wasn't systemetic. Data shows that 5 is a reasonable number. 
```{r}
# descriptive on the number of goals participants claimed to have prior to listing them 
describe(goalDf$total_goal)
hist(goalDf$total_goal)

# number of people who claimed having more than 5 goals
length(goalDf$total_goal[goalDf$total_goal>5])

# descriptive on the number of goals participants actual listed
describe(goalDf$listNum)
hist(goalDf$listNum)

# number of people who listed 1 goal
length(goalDf$listNum[goalDf$listNum == 1])

# descriptvie on the differences between claimed number and actual number
diffNum <- goalDf$total_goal - goalDf$listNum
describe(diffNum)
hist(diffNum)

# number of people who listed more goals than they claimed 
length(diffNum[diffNum <0])
diffNum[diffNum <0]

# number of people who listed less goals than they claimed
length(diffNum[diffNum >0])
diffNum[diffNum >0]
```

# Goal representation ratings

### Data transformation 
```{r}
# transform 99 & 999 to NAs
goalDf_long <- goalDf_long %>% 
  mutate(rating = replace(rating, rating == 99 | rating == 999, NA))
```

### Descriptive stats
From this, we can see that our subject tend to report goals that are attractive, attainable and intrinsically motivated. Subjects are highly committed even though their current progress are widely distributed. Most variables are positively skewed. Attractive_achievement and approach_avoidance_R are highly skewed and lack of variance. 
```{r}
# descriptive stats for each dimension 
dimensionDf <- goalDf_long %>%
  dplyr::select(dimension, rating) %>%
  group_by(dimension) %>%
  summarize(mean = mean(rating, na.rm = TRUE),
            sd = sd(rating, na.rm = TRUE), 
            n = n(),
            min = min(rating, na.rm = TRUE),
            max = max(rating, na.rm = TRUE),
            skew = skew(rating, na.rm = T), 
            kurtosi = kurtosi(rating, na.rm = T)
            )
# order based on their skewness 
kable(dimensionDf[order(dimensionDf$skew),])
```

```{r fig.height=14, fig.width=14}
# histograme for each dimension
goalDf_long %>%
  ggplot(aes(x = rating)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6) +
    facet_wrap(~dimension, nrow = 7)
```


### correlational matrix across all dimension
There's no dimension that doesn't correlate to anything else. However, there are some pairs that correlate too much: Importance & attractive_achiement & attractive_progress; ideal_motive & meaningness ;important & instrumentality; social_desirability & commonality; ought_motivation & external_motivation; attainment_maintenance & frequency 
```{r fig.height=20, fig.width=20}
# transform the long format to short format
goalDf_wide <- goalDf_long %>% spread (dimension, rating)

# generate a correlational matrix
corrM_all <- goalDf_wide %>% 
  dplyr :: select(affordance:visibility) %>% 
  cor(use = "pairwise.complete.obs")

# visualization
corrplot(corrM_all, method = "circle",number.cex = .7, order = "AOE", addCoef.col = "black",type = "upper")

```

### Variance Partition
94% of the variance was attributed to within subject differences. 
```{r}
# subset the long format dataset for only the 30 dimension
goal_striving <- c("initial_time_R", "advancement", "urgency", "effort", "commitment")
goalDf_R_long <- goalDf_long[!goalDf_long$dimension %in% goal_striving,]

# generate a multilevel model with subject as the random intercept
mlm <-lmer(rating ~ dimension + (1|id), data = goalDf_R_long)

# calculate the variance partition coefficient and transform to ICC
VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(icc=vcov/sum(vcov)) %>%
  dplyr :: select(grp, icc)
```

# Factor analysis

### EFA -- Raw rating score

27 varialbes are included. Goal preogress variables and ordinal rating variables are not included: temporal_duration & end_state_specificity and frequency
```{r}
# Exclude the 5 dimensions related to goal striving progress
goalDf_R_wide <- goalDf_wide[,!names(goalDf_wide) %in% goal_striving]

# Exclude ordinal variables: temporal_duration & end_state_specificity and frequency and other columns with irrelevent data
goal_ordinal <- c("temporal_duration", "end_state_specificity_R", "frequency")
goalDf_EFA <- goalDf_R_wide[,!names(goalDf_R_wide) %in% goal_ordinal]
goalDf_EFA <- subset(goalDf_EFA, select = affordance : visibility)

# Generate a correlational matrix 
corrM_raw <- cor(goalDf_EFA, use = "pairwise")

#corrplot(corrM_EFA, method = "circle",number.cex = .7, order = "AOE", addCoef.col = "black",type = "upper")
```

Determine the number of factors
```{r}
# use Very Simple Structure criterion
vss(corrM_raw, n = 10, rotate = "promax", diagonal = FALSE, fm = "minres", 
n.obs=NULL,plot=TRUE,title="Very Simple Structure",use="pairwise",cor="cor")

#nfactors(corrM_raw,n=10,rotate="promax",diagonal=FALSE,fm="minres",n.obs=NULL,
#             title="Number of Factors",pch=16,use="pairwise", cor="cor")


# use Scree plot & parallel analysis
ev <- eigen(corrM_raw)
ap <- parallel(subject=nrow(goalDf_EFA),var=ncol(goalDf_EFA),
  rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
plotnScree(nS)
```

vss suggest 4 factors, parallel suggested 6 factors and eigenvalues suggested 8 factors. we decided to use 6 factors because it fits better and the factors make sense
```{r}
# extract 4 factors
fa_raw_4 <-fa(r=corrM_raw, nfactors=4, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_raw_4)

# extract 6 factors
fa_raw_6 <-fa(r=corrM_raw, nfactors=6, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_raw_6)

# extract 8 factors
fa_raw_8 <-fa(r=corrM_raw, nfactors=8, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_raw_8)
```

visualization & interfactor correlation
```{r }
# factor diagram
fa.diagram(fa_raw_6)

# interfactor correlation
corr_fa <- fa_raw_6$Phi %>% as.data.frame() %>% rename(commonality = MR4, visibility = MR1, ought_motivation = MR2, difficulty = MR5, measurability = MR3, approach_avoidance_R = MR6)
rownames(corr_fa) <- c("commonality", "importance", "ought_motivation", "difficulty", "measurability", "approach_avoidance_R")
corr_fa
```


### EFA -- centered (within-subject deviation score)
```{r fig.height=20, fig.width=20}
# transform the ratings to the deviation score to the with-in subject mean on a given dimension
id <- data.frame("id" = goalDf_R_wide$id)

goalDf_meanCenter <- goalDf_EFA %>% bind_cols(id) %>% 
  group_by(id) %>%
  mutate_at(vars(-group_cols()), funs(scale(., scale = F))) %>%
  subset(select = -id)

# generate a correlational matrix 
corrM_meanCenter <- cor(goalDf_meanCenter,use = "pairwise")

# visualization 
corrplot(corrM_meanCenter, method = "circle",number.cex = .7, order = "AOE", addCoef.col = "black",type = "upper")
```

Determine number of factors vss suggested 4, parallel analysis suggested 5 and eigenevalues suggested 8
```{r}
# use Very Simple Structure criterion
vss(corrM_meanCenter, n = 10, rotate = "promax", diagonal = FALSE, fm = "minres", 
n.obs=NULL,plot=TRUE,title="Very Simple Structure",use="pairwise",cor="cor")

# Determine Number of Factors to Extract
ev <- eigen(cor(goalDf_meanCenter, use = "pairwise"))
ap <- parallel(subject=nrow(goalDf_meanCenter),var=ncol(goalDf_meanCenter),
  rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
plotnScree(nS)

# extract 8 factors from pairwise correlational matrix, using 
#factanal(factors = 5, covmat = corrM_meanCenter, rotation="varimax")

```

Extract factors
```{r}
fa_center_4 <- fa(r=corrM_meanCenter, nfactors=4, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_center_4)

fa_center_5 <- fa(r=corrM_meanCenter, nfactors=5, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_center_5)

fa_center_8 <- fa(r=corrM_meanCenter, nfactors=8, rotate="promax", SMC=FALSE, fm="minres")
fa.sort(fa_center_8)
```

visualization
```{r}
fa.diagram(fa_center_5)

# interfactor correlation
corr_fa <- fa_center_5$Phi %>% as.data.frame() %>% rename(importance = MR1, desirability = MR4, ought_motivation = MR2, attainability = MR3, measurability = MR5)
rownames(corr_fa) <- c("importance", "desirability", "ought_motivation", "attainability", "measurability")
corr_fa
```

### weighted average score for each factor (based on fa_raw_6)
```{r}
# Generate factor score for each subject using regression based weight
factorScoreDF <-factor.scores(goalDf_EFA, fa_raw_6, Phi = NULL, method = "Thurstone",rho=NULL,impute="mean")$scores %>% as.data.frame()

# Rename columns using the name of the primary variable
colnames(factorScoreDF) <- c("commonality", "importance", "ought_motivation", "difficulty", "measurability", "approach_avoidance_R")

# add subject IDs
factorScoreDF$id <- goalDf_wide$id
```

### variation partition on factor scores

difficulty has the highest within-subjedct variation. Commonality & importance have the lowest
```{r}
# commonality 
mlm <-lmer(commonality ~ 1 + (1|id), data = factorScoreDF)

commonality <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(commonality=vcov/sum(vcov)) %>%
  dplyr :: select(commonality)

# importance 
mlm <-lmer(importance ~ 1 + (1|id), data = factorScoreDF)

importance <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(importance=vcov/sum(vcov)) %>%
  dplyr :: select(importance)

# ought_motivation 
mlm <-lmer(ought_motivation ~ 1 + (1|id), data = factorScoreDF)

ought_motivation <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(ought_motivation=vcov/sum(vcov)) %>%
  dplyr :: select(ought_motivation)

# difficulty 
mlm <-lmer(difficulty ~ 1 + (1|id), data = factorScoreDF)

difficulty <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(difficulty=vcov/sum(vcov)) %>%
  dplyr :: select(difficulty)

# measurability 
mlm <-lmer(measurability ~ 1 + (1|id), data = factorScoreDF)

measurability <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(measurability=vcov/sum(vcov)) %>%
  dplyr :: select(measurability)

# approach_avoidance_R 
mlm <-lmer(approach_avoidance_R ~ 1 + (1|id), data = factorScoreDF)

approach_avoidance_R <- VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(approach_avoidance_R=vcov/sum(vcov)) %>%
  dplyr :: select(approach_avoidance_R)

# combine the outputs into one data frame
factorScore_icc <- data.frame("variation" = c("between subject", "within subject"))
factorScore_icc <- bind_cols(factorScore_icc, commonality, importance, ought_motivation, difficulty, measurability, approach_avoidance_R)
```

### goal level correlation between factor scores and goal pursuit variables

```{r fig.height=10, fig.width=10}
# combine goal level data
goalLevelDf <- cbind(factorScoreDF[,-7], goalDf_wide[,c("advancement", "commitment", "effort", "urgency", "initial_time_R")])

# generate a correlational matrix 
corrM_goalLevel <- cor(goalLevelDf,use = "pairwise")

# visualization 
corrplot(corrM_goalLevel, method = "circle",number.cex = .7, addCoef.col = "black",type = "upper")
```

### subject level aggregated factor scores
```{r}
# calculate the average score for each factor within each subject
subjectDf_factor <- aggregate(. ~ id, data = factorScoreDF, mean)

# calculate the sd for each factor wihtin each subject
subjectDf_factor_sd <- aggregate(. ~ id, data = factorScoreDF, sd)
```

# Euclidean distance

### use factor
use factor scores to calculate pairwise euclidean distance per subject
```{r}
# split the dataset by IDs
splitDf <- split( factorScoreDF, f = factorScoreDF$id)

# apply the distance function to each subject
output <- lapply(splitDf, function(x) dist(x,method = "euclidean", diag = FALSE, upper = FALSE))

# extract distance values
distance <- lapply(output, function(x) unlist(as.vector(x)))
distance <- unlist(distance)

# extract the number of pairs per subject
pairNum <- lapply(output, function(x) length(as.vector(x)))
pairNum <- unlist(pairNum)
mean(pairNum)

# generate a pairwise data frame
id <- unique(factorScoreDF$id)
id_pair <- unlist(mapply(rep, id, pairNum))

# pairId <- unlist(mapply(seq,1,pairNum))

pairDf <- data.frame("subject_id" = id_pair,
                     "distance" = distance)

# calculate mean distance per subject
distDf_perSub <- pairDf %>%
  group_by(subject_id) %>%
  mutate(distMean = mean(distance),
         distSd = sd(distance)) %>%
  dplyr :: select(-distance)

distDf_perSub <- aggregate(cbind(distMean,distSd) ~ subject_id, data = distDf_perSub, mean)

```

### descriptive of pairwise distance
```{r}
# descriptive of each pairwise distance
describe(pairDf$distance)

pairDf %>% ggplot(aes(x = distance)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)

# descriptive of subject level distance (raw score)
describe(distDf_perSub$distMean)

distDf_perSub %>% ggplot(aes(x = distMean)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)

# descriptive of subject level distance (variance score)
describe(distDf_perSub$distSd)

distDf_perSub %>% ggplot(aes(x = distSd)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)

```

### variance partition
```{r}
# generate a multilevel model with subject as the random intercept
mlm <-lmer(distance ~ 1 + (1|subject_id), data = pairDf)

# calculate the variance partition coefficient and transform to ICC
VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(icc=vcov/sum(vcov)) %>%
  dplyr :: select(grp, icc)
```

### use all variables 
```{r}
# split the dataset by IDs
goalDf_EFA$id <- goalDf_wide$id
splitDf <- split( goalDf_EFA, f = goalDf_EFA$id )

# apply the distance function to each subject
output <- lapply(splitDf, function(x) dist(x,method = "euclidean", diag = FALSE, upper = FALSE))

# extract distance values
distance_raw <- lapply(output, function(x) unlist(as.vector(x)))
distance_raw <- unlist(distance_raw)

# extract the number of pairs per subject
pairNum <- lapply(output, function(x) length(as.vector(x)))
pairNum <- unlist(pairNum)
mean(pairNum)

# generate a pairwise data frame
id <- unique(factorScoreDF$id)
id_pair <- unlist(mapply(rep, id, pairNum))

# pairId <- unlist(mapply(seq,1,pairNum))

pairDf_raw <- data.frame("subject_id" = id_pair,
                     "distance_raw" = distance_raw)

# calculate mean distance per subject
distDf_perSub_raw <- pairDf_raw %>%
  group_by(subject_id) %>%
  mutate(meanDist_raw = mean(distance_raw)) %>%
  dplyr :: select(-distance_raw)

distDf_perSub_raw <- aggregate(meanDist_raw ~ subject_id, data = distDf_perSub_raw, mean)
```

### variance partition
```{r}
# generate a multilevel model with subject as the random intercept
mlm <-lmer(distance_raw ~ 1 + (1|subject_id), data = pairDf_raw)

# calculate the variance partition coefficient and transform to ICC
VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(icc=vcov/sum(vcov)) %>%
  dplyr :: select(grp, icc)
```

### subject level correlations (with raw factor scores)
```{r fig.height=20, fig.width=20}
# combine subject level data frame (individual differences, mean factor scores, mean distance)
subjectDf <- left_join(subjectDf_factor, distDf_perSub, by = c("id" = "subject_id"))
subjectDf <- left_join(subjectDf, indivDiffDf, by = "id")

# generate a correlational matrix 
corrM_subjectRaw <- cor(subjectDf[,-1],use = "pairwise")

# visualization 
corrplot(corrM_subjectRaw, method = "circle",number.cex = .7, addCoef.col = "black",type = "upper")
```

### subject level correlations (with variance)

```{r fig.height=20, fig.width=20}
# combine subject level data frame (individual differences, variance in factor scores, mean and sd in distance)
subjectDf_sd <- left_join(subjectDf_factor_sd, distDf_perSub, by = c("id" = "subject_id"))
subjectDf_sd <- left_join(subjectDf_sd, indivDiffDf, by = "id")

# generate a correlational matrix 
corrM_subjectSd <- cor(subjectDf_sd[,-1],use = "pairwise")

# visualization 
corrplot(corrM_subjectSd, method = "circle",number.cex = .7, addCoef.col = "black",type = "upper")
```

```{r}
a <-lmer(rating ~ 1 + (1| id/goal), data = goalDf_long)
summary(a)
```

# Mahalanobis distance

### pair-wise distance
Use the covariance matrix across all goal ratings to calculate pair-wise distance
```{r}
# set a function for calculating pairwise distiance
mahalanobisFun <- function(df, cov) { 
  MD <- combn(nrow(df), 2, function(x) mahalanobis(as.matrix(df[x[1],]), center = as.matrix(df[x[2],]), cov = cov))
  return(tryCatch(MD, error=function(e) NULL))
}
```

excluded 2 subjects who only have 1 goal
```{r}
# tryCatch(testFunctionWihoutTryCatch(x), error=function(e) NULL)

# exclude subjects with only one goal
id_oneGoal <- goalDf_wide$id[goalDf_wide$listNum ==1]
factorScoreDF_clean <- factorScoreDF %>% filter(!id %in% id_oneGoal)

# split the dataset by IDs and then get rid off the ID column
splitDf <- split( factorScoreDF_clean, f = factorScoreDF_clean$id)
#splitDf <- split( factorScoreDF, f = factorScoreDF$id)
splitDf <- lapply(splitDf, function(x) subset(x, select = -id))

# get the covariance matrix on factor scores across all goals
factor_cov <- cov(subset(factorScoreDF_clean, select = -id))
#factor_cov <- cov(subset(factorScoreDF, select = -id))

# apply the distance function to each subject
output <- lapply(splitDf, function(x) mahalanobisFun(x, factor_cov))
#output <- lapply(splitDf, function(x) tryCatch(mahalanobisFun(x, factor_cov), error = function(e) NULL))

# extract distance values
distance_M <- unlist(output)

# extract the number of pairs per subject
pairNum <- lapply(output, function(x) length(as.vector(x)))
pairNum <- unlist(pairNum)
mean(pairNum)

# generate a pairwise data frame
id <- unique(factorScoreDF_clean$id)
id_pair <- unlist(mapply(rep, id, pairNum))

# pairId <- unlist(mapply(seq,1,pairNum))

pairDf_M <- data.frame("subject_id" = id_pair,
                     "distance_M" = distance_M)
```

```{r}
# descriptive of all pairwise distance
describe(pairDf_M$distance_M)

pairDf_M %>% ggplot(aes(x = distance_M)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)
```

### variance partition
```{r}
# generate a multilevel model with subject as the random intercept
mlm <-lmer(distance_M ~ 1 + (1|subject_id), data = pairDf_M)

# calculate the variance partition coefficient and transform to ICC
VarCorr(mlm) %>%
  as_data_frame() %>%
  mutate(icc=vcov/sum(vcov)) %>%
  dplyr :: select(grp, icc)
```

### subject-level distance (mean & sd)
```{r}
# calculate mean distance per subject
distDf_perSub_M <- pairDf_M %>%
  group_by(subject_id) %>%
  mutate(distMean = mean(distance_M),
         distSd = sd(distance_M)) %>%
  dplyr :: select(-distance_M)

distDf_perSub_M <- aggregate(cbind(distMean,distSd) ~ subject_id, data = distDf_perSub_M, mean)

# descriptive of subject-level mean distance
describe(distDf_perSub_M$distMean)

distDf_perSub_M %>% ggplot(aes(x = distMean)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)

# descriptive of subject-level distance sd
describe(distDf_perSub_M$distSd)

distDf_perSub_M %>% ggplot(aes(x = distSd)) +
    geom_histogram(fill   = "orange",
                   colour = "black",
                   alpha  = .6)
```

### subject level correlations (with raw factor scores)
```{r fig.height=20, fig.width=20}
# combine subject level data frame (individual differences, mean factor scores, mean distance)
subjectDf <- left_join(subjectDf_factor, distDf_perSub_M, by = c("id" = "subject_id"))
subjectDf <- left_join(subjectDf, indivDiffDf, by = "id")

# generate a correlational matrix 
corrM_subjectRaw <- cor(subjectDf[,-1],use = "pairwise")

# visualization 
corrplot(corrM_subjectRaw, method = "circle",number.cex = .7, addCoef.col = "black",type = "upper")
```

### subject level correlations (with variance)
```{r fig.height=20, fig.width=20}
# combine subject level data frame (individual differences, variance in factor scores, mean and sd in distance)
subjectDf_sd <- left_join(subjectDf_factor_sd, distDf_perSub_M, by = c("id" = "subject_id"))
subjectDf_sd <- left_join(subjectDf_sd, indivDiffDf, by = "id")

# generate a correlational matrix 
corrM_subjectSd <- cor(subjectDf_sd[,-1],use = "pairwise")

# visualization 
corrplot(corrM_subjectSd, method = "circle",number.cex = .7, addCoef.col = "black",type = "upper")
```